Type-Safe Client-Server Communication in Web Applications

### 2014-02-27 paper.tex
Add subsection about server-side state.

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, both client and server code may lift
        general IO computations, allowing for safe IO code reuse within the
        confines of the client or server designated code sections.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item Our model takes the view that the client side is the main driver when
        developing web applications and accordingly delegates the server to
        the role of a computational and/or storage resource, tasked with
        servicing client requests rather than driving the program.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no compiler support.
\end{itemize}

\subsection{A first example}

@@

Like \lstinline!runClient!, \lstinline!runApp! does not return, and is intended
to be the only computation executed by \lstinline!main!. In fact, performing
non-deterministic IO and allowing \lstinline!runApp! or its arguments to
capture the result may lead to unpredictable behavior. A remedy for this is
discussed in section \ref{sec:discussion}.

@@

@@

\section{Discussion and related work}\label{sec:discussion}

@@

\subsection{Limitations}

@@

\paragraph{Preventing divergence} While our solution completely safe from the
invocation point of the \lstinline!runApp! function onward, capturing
IO-dependent variables introduced before this invocation may lead to the client
and server execution paths diverging. Consider the following example:

>>>>

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, both client and server code may lift
        general IO computations, allowing for safe IO code reuse within the
        confines of the client or server designated code sections.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item Our model takes the view that the client side is the main driver when
        developing web applications and accordingly delegates the server to
        the role of a computational and/or storage resource, tasked with
        servicing client requests rather than driving the program.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no compiler support. Programs
        are compiled twice; once with Haste and once with GHC, to produce the
        final client and server side code respectively.
\end{itemize}

\subsection{A first example}\label{sec:helloserver}

@@

Like \lstinline!runClient!, \lstinline!runApp! does not return, and is intended
to be the only computation executed by \lstinline!main!. In fact, performing
non-deterministic IO and allowing \lstinline!runApp! or its arguments to
capture the result may lead to unpredictable behavior. The \lstinline!appMain!
function should for all intents and purposes be regarded as the program's entry
point; a non-intrusive method of enforcing this using the type system is
described in section \ref{sec:limitations}, but is not used in the examples
given here as it relies on the use of the GHC plugin system.

@@

\subsection{Using server-side state}

While the Hello Server example illustrates how client-server communication is
handled, most web applications need to keep some server-side state as well.
At first glance, this would seem to be a trivial problem; can we not just
create our state holding elements in \lstinline!main! as usual?
Figure \ref{lst:bad-state} describes an attempt at a solution using this idea.

\begin{listingfloat}
\begin{code}
main = do
  visitorref <- newIORef 0
  runApp config $ do
    count <- export . liftIO $ do
      atomicModifyIORef visitorref (\v -> (v+1, v+1))

    runClient $ do
      visitors <- onServer count
      alert ("Your are visitor #" ++ show visitors)
\end{code}
\caption{Server-side state: a misguided first attempt}
\label{lst:bad-state}
\end{listingfloat}

While this may at first seem to solve the problem of server-side state, this
approach raises a number of questions. Does \lstinline!visitorref! live on the
client or the server, or perhaps on both? What happens if our state holding
element is initialized to a non-deterministic value, and what if we assign it
a value on the server and then attempt to read that value on the client?
Recall that, as explained in section \ref{sec:helloserver}, the function
passed to \lstinline!runApp! should for all intents and purposes be regarded as
the program's entry point.

To allow the use of server state while treating the final argument to
\lstinline!runApp! (from here on always referred to as \lstinline!appMain!)
as the program's entry point, we may want to allow arbitrary IO computations to
be lifted into the \lstinline!App! monad:

\begin{code}
main = runApp config $ do
  visitorref <- liftIO $ newIORef 0
  ...
\end{code}

However, this does not answer the question of where \lstinline!visitorref!
lives or where any side effects caused by its creation will take place.
To solve this problem once and for all, we need to introduce a way to lift
arbitrary IO computations, but ensure that said computations are executed on
the server and nowhere else. This is accomplished using a more restricted
version of \lstinline!liftIO!:

\begin{code}
liftServerIO :: IO a -> App (Useless a)
\end{code}

In contrast to \lstinline!liftIO!, the result value of \lstinline!liftServerIO!
is wrapped in the \lstinline!Useless! type. As the name implies, this type is
completely opaque and thus useless unless you can somehow extract its contained
value, something which is only possible using a single function:

\begin{code}
mkUseful :: Useless a -> Server a
\end{code}

Therein lies the secret to restricting \lstinline!liftServerIO! to the
server: the only function capable of turning a \lstinline!Useless! value into
a useful one lives in the server monad. Any client side code is thus free to
completely ignore executing computations lifted using \lstinline!liftServerIO!;
since the result of such a computation is never observable on the client,
the client has no obligation to even produce such a value.
Figure \ref{lst:good-state} shows how to make proper use of server-side state.

\begin{listingfloat}
\begin{code}
main =
  runApp config $ do
    uselessref <- liftServerIO $ newIORef 0

    count <- export $ do
      r <- mkUseful uselessref
      liftIO $ atomicModifyIORef r (\v -> (v+1, v+1))

    runClient $ do
      visitors <- onServer count
      alert ("Your are visitor #" ++ show visitors)
\end{code}
\caption{Server-side state: doing it properly}
\label{lst:good-state}
\end{listingfloat}

@@

\section{Discussion and related work}

@@

\subsection{Limitations}
\label{sec:limitations}

@@

\paragraph{Preventing divergence} While our solution completely safe from the
invocation point of the \lstinline!runApp! function onward, capturing
IO-dependent variables introduced before this invocation may lead to the client
and server execution paths diverging. Recall that our solution does not modify
either of the compilers used. As such, both will produce code to execute all
computations called by \lstinline!main!. Consider the following example:
